Automatically Proving Equivalence by Type-Safe Reflection

### 2017-03-30 paper-CICM-17/content/1-introduction.tex
some fixes

Proofs assistants like Coq~\cite{BertotC04}, Agda(ref) and Idris~\cite{brady2013idris} are based on Intensional Type Theories, that contain two notions of equality. One is the equality that can be manipulated in the language, called the propositional equality. This equality corresponds to the equality found in mathematics : this is a proposition that can be assumed, negated, proved or disproved. Since in type theory, propositions are seen as types~\cite{How80}, the
proposition that two elements $x$ and $y$ are equal corresponds to a type.
If $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$. Thus, $Id$ is a type family (parametrised by the type $a$) indexed over two elements of $a$, giving $Id (a:Type)\ :\ a \rightarrow a \rightarrow Type$. For
convenience, we write $(Id_a\ x\ y)$ as $(x\ =_a\ y)$.

@@

The only way for $(Id_a\ x\ y)$ to be inhabited is therefore that $x$ should be
(by definition) $y$. In this case, the constructor $Refl$ helps to create a
proof of this equality : $(Refl\ x)$ is precisely the proof which says that
$x=_ax$.

@@

then we have : \\
$T\_ind : \forall P:T \rightarrow Type,\ (P\ T0) \rightarrow (\forall t:T,\ P\ t \rightarrow P\ (T1\ t)) \rightarrow (\forall t:T,\ P\ t)$.

The principles of induction can be used for proving any proposition $P$, so they also work for the propositional equality. For example, we can prove that $n+0 = n$ for all $n$ by induction on the $Nat\ n$, even if $n+0 \not\equiv n$ with the usual definition of $+$, recursive on its first argument. That means that the axiom of induction has made the type $Id_{a}(x,y)$ bigger than it looks like : it contains terms that are not generated by $Refl$ (the only constructor of $Id_{a}(x,y)$), but that are added by the inductive principles. There are therefore things which are provably equal which unfortunately are not definitionally equal after unfolding all the definitions and after full reductions. The activity of proving equalities is therefore in these theories something which isn't automatically decidable by the type-checker in the very general case.

@@

We revisit an example from a previous work of Edwin Brady~\cite{DBLP:conf/plpv/Brady13} which shows how proof obligations may arise when a type is indexed by natural numbers. In this development, the goal is to implement a formally verified library of binary numbers. In order to ensure the correctness of the methods that will be developed, we define the types $Bit$ and $Binary$ as indexed over the value they represent (expressed as a natural number), following a correct-by-construction approach.

@@

This function produces a dependent pair. The first component is the value $bX$ (expressed as a natural number) representing the first bit produced (therefore of type $Bit\ bX$), and the second component is another dependent pair. This second dependent pair is composed of the value $bY$ representing the second bit produced, and of a triple containing the two bits produced and a proof that these two bits correctly encode the result of the addition.
For example, on the fifth line, which corresponds to the computation $1_2 + 0_2 + 0_2 = (01)_2$, the function produces the bit $b0$ (encoding $0_2$) and the bit $b1$ (encoding $1_2$), and a proof that $1 + 0 + 0 = 1 + (2*0)$.

We finally define the function $adc$ that adds two binary numbers and a carry flag. This addition works for two binary numbers expressed with the same number of bits, and it produces a result with one more bit. This result represents the sum of the three concrete values of the inputs. Therefore, we want to write :

@@

Unfortunately, this straightforward definition is rejected by Idris' type-checker because the types mismatch for both of the two patterns. The result of the first line $adc\ zero\ zero\ carry$ is expected to have the type $Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)$ but we provide a term of type $Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0)))$.
The problem is even worse for the second case where the expected index is \\
$(plus\ (plus\ c\ (plus\ bit2\ (plus\ v1\ (plus\ v1\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))$
while we're trying to provide a term indexed over \\
$plus\ vLsb\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ 0))$.
We know that the definition of $adc$ we have given is correct, and it has \emph{provably} the expected type, but it does not have it \emph{immediately} or \emph{by reduction} : after full reductions\footnote{In dependently typed theories, the presence of types predicated on values creates the need of using reductions also inside of types}, the expected and provided types are still different, and Idris can not figure out a proof of compatibility on its own. 
If we use provisional definitions[ref], the previous definition can be accepted by the system, but we now have two proof obligations $adc\_lemma\_1$ and $adc\_lemma\_2$ demanding proofs of equality between the two types, and they have to be provided by hand --here with a proof script-- :

@@

As we see, this kind of proof consists of a potentially long sequence of rewriting steps, where each step uses one of the available properties: neutral element, commutativity, associativity. Without some specific automation, this sequence of rewriting must be done by the programmer.
This is time consuming and more importantly, just a small change in the left or the right hand side of the equality may invalidate the proof. That means that a minor change in our datatypes, or in the definition of $addBit$ or $adc$ will require us to replace this proof by another one fulfilling the same task.
These proofs become the everyday routine in any dependently-typed language, as soon as we index types over values in order to capture some logical properties.

The proof that we have written by hand only uses the existence of a neutral element, and the associativity and commutativity of $+$ on $Nat$. Thus, we're rewriting a term by using the properties of a commutative monoid. We would like to let a generic prover for commutative monoid find a proof on its own.

>>>>

Proofs assistants like Coq~\cite{BertotC04}, Agda~\cite{DBLP:conf/tldi/Norell09} and Idris~\cite{brady2013idris} are based on Intensional Type Theories, that contain two notions of equality. One is the equality that can be manipulated in the language, called the propositional equality. This equality corresponds to the equality found in mathematics : this is a proposition that can be assumed, negated, proved or disproved. Since in type theory, propositions are seen as types~\cite{How80}, the
proposition that two elements $x$ and $y$ are equal corresponds to a type.
If $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$. Thus, $Id$ is a type family (parametrised by the type $a$) indexed over two elements of $a$, giving $Id (a:Type)\ :\ a \rightarrow a \rightarrow Type$. For
convenience, we write $(Id_a\ x\ y)$ as $(x\ =_a\ y)$.

@@

The only way for $(Id_a\ x\ y)$ to be inhabited is therefore that $x$ should be
(by definition) $y$. In this case, the constructor \code{Refl} helps to create a
proof of this equality : \code{(Refl\ x)} is precisely the proof which says that
$x=_ax$.

@@

then we have the following induction principle for $T$: \\
$T\_ind : \forall P:T \rightarrow Type,\ (P\ T0) \rightarrow (\forall t:T,\ P\ t \rightarrow P\ (T1\ t)) \rightarrow (\forall t:T,\ P\ t)$.

The principles of induction can be used for proving any proposition $P$, so they also work for the propositional equality. For example, we can prove that $n+0 = n$ for all $n$ by induction on the $Nat\ n$, even if $n+0 \not\equiv n$ with the usual definition of $+$, recursive on its first argument. That means that the axiom of induction has made the type $Id_{a}(x,y)$ bigger than it looks like : it contains terms that are not generated by \code{Refl} (the only constructor of $Id_{a}(x,y)$), but that are added by the inductive principles. There are therefore things which are provably equal which unfortunately are not definitionally equal after unfolding all the definitions and after full reductions. The activity of proving equalities is therefore in these theories something which isn't automatically decidable by the type-checker in the very general case.

@@

We revisit an example from a previous work of Edwin Brady~\cite{DBLP:conf/plpv/Brady13} which shows how proof obligations may arise when a type is indexed by natural numbers. In this development, the goal is to implement a formally verified library of binary numbers. In order to ensure the correctness of the methods that will be developed, we define the types \code{Bit} and \code{Binary} as indexed over the value they represent (expressed as a natural number), following a correct-by-construction approach.

@@

This function produces a dependent pair. The first component is the value \code{bX} (expressed as a natural number) representing the first bit produced (therefore of type \code{Bit\ bX}), and the second component is another dependent pair. This second dependent pair is composed of the value \code{bY} representing the second bit produced, and of a triple containing the two bits produced and a proof that these two bits correctly encode the result of the addition.
For example, on the fifth line, which corresponds to the computation $1_2 + 0_2 + 0_2 = (01)_2$, the function produces the bit \code{b0} (encoding $0_2$) and the bit \code{b1} (encoding $1_2$), and a proof that $1 + 0 + 0 = 1 + (2*0)$.

We finally define the function \code{adc} that adds two binary numbers and a carry flag. This addition works for two binary numbers expressed with the same number of bits, and it produces a result with one more bit. This result represents the sum of the three concrete values of the inputs. Therefore, we want to write :

@@

Unfortunately, this straightforward definition is rejected by Idris' type-checker because the types mismatch for both of the two patterns. The result of the first line \code{adc zero zero carry} is expected to have the type \codeL{Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)} but we provide a term of type \code{Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0)))}.
The problem is even worse for the second case where the expected index is \codeL{(plus\ (plus\ c\ (plus\ bit2\ (plus\ v1\ (plus\ v1\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))}
while we're trying to provide a term indexed over
\codeL{plus\ vLsb\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ 0))}.
We know that the definition of \code{adc} we have given is correct, and it has \emph{provably} the expected type, but it does not have it \emph{immediately} or \emph{by reduction} : after full reductions\footnote{In dependently typed theories, the presence of types predicated on values creates the need of using reductions also inside of types}, the expected and provided types are still different, and Idris can not figure out a proof of compatibility on its own. 
If we use provisional definitions[ref], the previous definition can be accepted by the system, but we now have two proof obligations \code{adc\_lemma\_1} and \code{adc\_lemma\_2} demanding proofs of equality between the two types, and they have to be provided by hand --here with a proof script-- :

@@

As we see, this kind of proof consists of a potentially long sequence of rewriting steps, where each step uses one of the available properties: neutral element, commutativity, associativity. Without some specific automation, this sequence of rewriting must be done by the programmer.
This is time consuming and more importantly, just a small change in the left or the right hand side of the equality may invalidate the proof. That means that a minor change in our datatypes, or in the definition of \code{addBit} or \code{adc} will require us to replace this proof by another one fulfilling the same task.
These proofs become the everyday routine in any dependently-typed language, as soon as we index types over values in order to capture some logical properties.

The proof that we have written by hand only uses the existence of a neutral element, and the associativity and commutativity of $+$ on \code{Nat}. Thus, we're rewriting a term by using the properties of a commutative monoid. We would like to let a generic prover for commutative monoid find a proof on its own.

### 2017-03-30 paper-CICM-17/content/3-hierarchyOfTactics.tex
some fixes

What we want to build is a prover able to prover not only equalities for $Nat$, $List$,or nay specific type but for various datatypes, and for various properties. With some abstraction, we can solve the problem of generating proof of equality for many datatypes like natural numbers, lists, matrix, and many others \emph{all at once} by implementing a very generic hierarchy of provers that can prove equalities of terms for Semi-Groups, Monoids, Commutative-Monoids, Groups, Commutative-Groups, Semi-Rings and Rings. These tactics will be usable on any type that satisfies the properties of one of these algebraic structure. The properties of a given algebraic algebraic structure will be expressed in a corresponding interface (sometimes called typeclass in other progamming languages). This interface will extend the interface from which it naturally inherits. For example, the $Group$ interface will extend the $Monoid$ one. We will end up with a hierarchy of interfaces and will write one tactic for each of these interface. All the tactics that we will implement (the ring prover, the group prover, etc) will be able to work on any type, being given that an implementation of the corresponding interface is provided.

@@

With some reasonable additional effort, we could produce a collection of tactics for proving equivalences instead of only equality. The machinery would be very similar, and if we do the right abstraction from the start, we could gain one more degree of genericity, with the freedom of choosing the equivalence relation (which can of course be the usual equality). We want to give to the user the possibility to define his own equivalence notion, as long as he's able to provide the properties of the algebraic structure he wants to use. Let's call $c$ the \emph{carrier} type, ie, the type on which we want to prove equivalences. The equivalence relation on $c$ has the following profile $(\simeq) : c \rightarrow c \rightarrow Type$\footnote{This 'Type' would be a 'Prop' in systems --like Coq-- that make a distinction between the world of computations and the world of logical statements}, and has to be accompanied by the usual properties of reflexivity, symmetry and transitivity.

All of our tactics will require to have a way of testing this equivalence between elements of the underlying set, that is to say, a way to test equivalence between constants. For this reason, we define a notion of Set\footnote{This notion of Set isn't a formalisation of sets and their elements, but is only a way to talk about the carrier type and an equivalence relation, sometimes called Setoid}, which only requires the definition of the equivalence relation --accompanied with the proofs that this is indeed an equivalence relation-- and this equivalence test $set\_eq$. All the algebraic structures will later extend this interface :

@@

If the user wants to prove propositional equalities, then he will simply instantiate $(\simeq)$ with the built-in $(=)$ during the definition of the $Set$ implementation.
Note that $(\simeq)$ is only weakly decidable in the sense that $set\_eq$ only produces a proof when the two elements are equivalent, but it doesn't produce a proof of dis-equivalence when they are different -- instead, it simply produces the value $Nothing$--. That's natural, as we want to generate proof of equivalence, and not to generate counter examples for proving dis-equivalence, which is another problem.

Obviously, there is no tactic associated to $Set$, since we have no operations and no properties associated to this structure. Therefore, equivalences in a $Set$ are just the "syntactic equivalences", and they can simply be proven with $refl$\footnote{$refl$ should not be confused with $Refl$, the only constructor of $=$. But of course, when $(\simeq)$ is instantiated with the equality $=$, the instantiation of $refl$ will be $Refl$. Therefore, $refl$ of the interface $Set$ is a generalisation of the most specialised $Refl$ for $=$}.

@@

Working with equivalences instead of equalities will bring one complication. These proofs of correctness that we will have to produce by hand will be slightly more complicated, as we won't be able to use idris' "rewrite" command, that enables to rewrite a subterm of a term by another one, provided that the two subterms are equal. More precisely, if we need to prove $P\ x$, and have a proof of equivalence $pr:x \simeq x'$, we can't use "rewrite pr" in order to transform the goal into $P\ x'$, as we would normally do if the available proof would be an equality $x=x'$. This is a classical problem of working within a setoid, and this problem can be slightly mitigated when the programming language offers a good support for rewriting terms in setoids. However, Idris isn't equipped with any automation for setoids, ans everything will have to be done by hand. For this purpose, we will use the following lemma :

$eq\_preserves\_eq : \{c:Type\} \rightarrow (Set\ c) \rightarrow (x:c) \rightarrow (y:c) \rightarrow (c1:c) \rightarrow (c2:c) \rightarrow (x \simeq c1) \rightarrow (y \simeq c2) \rightarrow (c1 \simeq c2) \rightarrow (x \simeq y)$.

\begin{proof}
We have $x \simeq c1$ and we have $c1 \simeq c2$, therefore we have $x \simeq c2$ by the use of the axiom of transitivity (of the interface $Set$). \

We also have $y \simeq c2$ and therefore we have $c2 \simeq y$ by the use of the symmetry axiom (of the interface $Set$). \

@@

The operations, constants and properties of each algebraic structure will be described in a dedicated interface. The first structure, completely trivial, is the magma. A magma is a structure built on top of $Set$, that just adds a $Plus$ operation, and no specific properties about it :

@@

This code means that a type $c$ (for $carrier$) is a Magma if it is already a Set (ie, it is equipped with the equivalence relation $\simeq$ and the equivalence test $set\_eq$), and if it has a $+$ operation.
In fact, there is an additional requirement that will apply for all operators (in this case, the $+$ operation), which is that they need to be "compatible" with the equivalence relation, which is expressed by the following axiom for $+$ : \\ $Plus\_preserves\_equiv : \{c1:c\} \rightarrow \{c2:c\} \rightarrow \{c1':c\} \rightarrow \{c2':c\} \rightarrow (c1 \simeq c1') \rightarrow (c2 \simeq c2') \rightarrow ((c1+c2) \simeq (c1'+c2'))$ \\ 
This requirement comes from the fact that we're dealing with any equivalence relation. The user is free to use his own equivalence relation, but it should be compatible with the operations that he is using.
As it was the case for $Set$, there is no tactic to write for $Magma$, because there is no property at all : all equivalences are again syntactic equivalences, and they can all be proven by $refl$.

A semi-group is a magma (ie, it still has a $Plus$ operation), but moreover it has the property of associativity for this operation.

@@

As example of magma, we can cite $Nat$ equipped with its operation of addition, $List$ and the operation of concatenation.

Now, the Monoid structure is a Semi-Group with the property of neutral element for a distinguished element called $Zero$.

@@

The hierarchy of interfaces continues with Commutative-Monoid, Group, Commutative Group, Semi-Ring and Ring, but we only show the Group interface for reasons of space :

@@

The notion of group uses two new operations ($Neg$ and $Minus$), but $Minus$ is not primitive and can be simplified with $+$ and $Neg$. The important property of a group is that every element $c1$ must admit $Neg\ c1$ as inverse element for the $+$ operation.

@@

When trying to prove an equivalence $x \simeq y$, the universally-quantified variables have been abstracted, and they are now part of the context. Our tatics will work by normalising both sides of the "potential equivalence" $x \simeq y$, and afterwards comparing them with a syntactic equality test. The problem is that such a normalisation function will need to do different treatments for a a variable, for a constant and for an operator. For this reason, we will need to work by reflection. There will be one type of reflected terms for each algebraic structure. The novelty is not the use of reflected terms (this has already been done, as discussed in section~\ref{sect:relatedWork}), but is the use of a type-safe reflection mechanism where we index the reflected terms by the concrete value that they represent. Each of these datatype is parametrised over a type $c$, which is the real type on which we want to prove equalities (the $carrier$ type). It is also indexed over an implementation of the corresponding interface for $c$ (we usually call it $p$, because it behaves as a $proof$ telling that the structure $c$ has the desired properties), indexed over a context of abstracted variables (a vector $\Gamma$ of $n$ elements of type $c$), and most important it is also indexed over a value of type $c$, which is precisely the concrete value being encoded.

A magma is only equipped with one operation $+$. Thus, we only have three concepts to express in order to reflect terms in a Magma : constants, variables, and additions.

@@

For an expression $e_x\ :\ ExprMa\ \Gamma\ x$, we will say that $e_x$ denotes (or encodes) the number $x$ in the context $\Gamma$.
When an expression is a variable $(VarMa\ \_\ \Gamma\ i)$, the denoted number is simply the corresponding variable in the context, ie, $(index\ i\ \Gamma)$.
The expression $(ConstMa\ \_\ \Gamma\ k)$ denotes the constant $k$ of type $c$. Finally, if $e_x$ is an expression encoding the concrete value $x$, and $e_y$ is an expression encoding the concrete value $y$, then the expression $PlusMa\ e_x\ e_y$ denotes the concrete value $(x + y)$. Because the reflected terms embed their corresponding inputs, they are guarantee to be faithful representations of them.

There is no additional operations in SemiGroup and Monoid, so the datatypes that reflect terms in these two structures will have exactly the same shape as the one for Magma that we have given above.
However, the one for $Group$ will introduce two new constructors for the $Neg$ and $Minus$ operations :

@@

The index of type $c$ (the real value encoded by an expression) is always expressed by using the lookup function $index$ and the available operations in the implementation $p$, which for a group are $+$, $Minus$ and $Neg$.

@@

We want to write the reduction function on a \emph{correct by construction} way, which means that no additional proof should be required after the definition of the function. Thus, $norm$ will produce the proof that the new expression freshly produced has the same interpretation as the original expression, and this will be made easier by the fact that all the datatypes for reflecting terms ($ExprSG$, $ExprG$, $ExprR$) are indexed over the real --concrete-- value : a term of type $Expr\ \Gamma\ x$ is the encoding of the concrete value $x$ in the context $\Gamma$.
Thus, for each algebraic structure, the type of the function $norm$ will look like this : \\
$norm :\ Expr\ \Gamma\ x\ \rightarrow\ (x'\ **\ (Expr\ \Gamma\ x',\ x\ \simeq\ x'))$ \\
Every function of normalisation $norm$ will produce a dependent pair : the new concrete value $x'$, and a pair made of an $Expr\ \Gamma\ x'$ which is the new encoded term indexed over the new concrete number we have just produced, and a proof that old and new concrete values $x$ and $x'$ are equivalent.

This function doesn't simply produce an $Expr\ \Gamma\ x$ (with an internal conversion of type) because the proof of $x \simeq x'$ is an essential component for building the desired proof of $x \simeq y$, which is the main problem that we are trying to solve.

We will explain how the tactic works by commenting the construction of the tactic for Group, but the other algebraic structures will be equipped similarly. The equivalence we are trying to prove is $x \simeq y$, where $x$ and $y$ are elements of the type $c$, which  simulates a group. The reflected term for $x$ will be denoted $e_x$, and this term will have the type $ExprG\ p\ \Gamma\ x$, which means that $e_x$ is guaranteed to be the encoding of $x$ because $e_x$ is precisely indexed over the real value $x$.
We've got similar things for $y$, which is encoded by $e_y$, and its type is indexed over the real value $y$.
Running the normalisation procedure on $e_x$ will produce the normal form $e_{x'}$ of type $ExprG\ p\ \Gamma\ x'$ and a proof $p_x$ of $x \simeq x'$ while running the normalisation procedure on $e_y$ will produce the normal form $e_{y'}$ of type $ExprG\ p\ \Gamma\ y'$ and a proof $p_y$ of $y \simeq y'$.
It is now enough to simply compare $e_{x'}$ and $e_{y'}$ using a standard syntactic equality test $exprG\_eq$ because these two expressions are now supposed to be in normal form.

>>>>

What we want to build is a prover able to prover not only equalities for \code{Nat}, \code{List}, or any specific type but for various datatypes, and for various properties. With some abstraction, we can solve the problem of generating proof of equality for many datatypes like natural numbers, lists, matrix, and many others \emph{all at once} by implementing a very generic hierarchy of provers that can prove equalities of terms for Semi-Groups, Monoids, Commutative-Monoids, Groups, Commutative-Groups, Semi-Rings and Rings. These tactics will be usable on any type that satisfies the properties of one of these algebraic structure. The properties of a given algebraic algebraic structure will be expressed in a corresponding interface (sometimes called typeclass in other progamming languages). This interface will extend the interface from which it naturally inherits. For example, the \code{Group} interface will extend the \code{Monoid} one. We will end up with a hierarchy of interfaces and will write one tactic for each of these interface. All the tactics that we will implement (the ring prover, the group prover, etc) will be able to work on any type, being given that an implementation of the corresponding interface is provided.

@@

With some reasonable additional effort, we could produce a collection of tactics for proving equivalences instead of only equality. The machinery would be very similar, and if we do the right abstraction from the start, we could gain one more degree of genericity, with the freedom of choosing the equivalence relation (which can of course be the usual equality). We want to give to the user the possibility to define his own equivalence notion, as long as he's able to provide the properties of the algebraic structure he wants to use. Let's call $c$ the \emph{carrier} type, ie, the type on which we want to prove equivalences. The equivalence relation on $c$ has the following profile $(\simeq) : \code{c} \rightarrow \code{c} \rightarrow \code{Type}$\footnote{This \code{Type} would be a \code{Prop} in systems --like Coq-- that make a distinction between the world of computations and the world of logical statements}, and has to be accompanied by the usual properties of reflexivity, symmetry and transitivity.

All of our tactics will require to have a way of testing this equivalence between elements of the underlying set, that is to say, a way to test equivalence between constants. For this reason, we define a notion of \code{Set}\footnote{This notion of set isn't a formalisation of sets and their elements, but is only a way to talk about the carrier type and an equivalence relation, sometimes called Setoid}, which only requires the definition of the equivalence relation --accompanied with the proofs that this is indeed an equivalence relation-- and this equivalence test \code{set\_eq}. All the interfaces representing the algebraic structures will later extend \code{Set} :

@@

If the user wants to prove propositional equalities, then he will simply instantiate $(\simeq)$ with the built-in $(=)$ during the definition of the \code{Set} implementation.
Note that $(\simeq)$ is only weakly decidable in the sense that \code{set\_eq} only produces a proof when the two elements are equivalent, but it doesn't produce a proof of dis-equivalence when they are different -- instead, it simply produces the value \code{Nothing}--. That is normal, as we want to generate proof of equivalence, and not to generate counter examples for proving dis-equivalence, which is another problem.

Obviously, there is no tactic associated to \code{Set}, since we have no operations and no properties associated to this structure. Therefore, equivalences in a \code{Set} are just the "syntactic equivalences", and they can simply be proven with \code{refl}\footnote{\code{refl} should not be confused with \code{Refl}, the only constructor of $=$. But of course, when $(\simeq)$ is instantiated with the equality $=$, the instantiation of \code{refl} will be \code{Refl}. Therefore, \code{refl} of the interface \code{Set} is a generalisation of the most specialised \code{Refl} for \code{=}}.

@@

Working with equivalences instead of equalities will bring one complication. These proofs of correctness that we will have to produce by hand will be slightly more complicated, as we won't be able to use idris' "rewrite" command, that enables to rewrite a subterm of a term by another one, provided that the two subterms are equal. More precisely, if we need to prove $P\ x$, and have a proof of equivalence $pr:x \simeq x'$, we can't use "rewrite pr" in order to transform the goal into $P\ x'$, as we would normally do if the available proof would be an equality $x=x'$. This is a classical problem of working within a setoid, and this problem can be slightly mitigated when the programming language offers a good support for rewriting terms in setoids. However, Idris isn't equipped with any automation for setoids, and everything will have to be done by hand. For this purpose, we will use the following lemma :

$\code{eq\_preserves\_eq} : \{c:Type\} \rightarrow (Set\ c) \rightarrow (x:c) \rightarrow (y:c) \rightarrow (c1:c) \rightarrow (c2:c) \rightarrow (x \simeq c1) \rightarrow (y \simeq c2) \rightarrow (c1 \simeq c2) \rightarrow (x \simeq y)$.

\begin{proof}
We have $x \simeq c1$ and we have $c1 \simeq c2$, therefore we have $x \simeq c2$ by the use of the axiom of transitivity (of the interface \code{Set}). \

We also have $y \simeq c2$ and therefore we have $c2 \simeq y$ by the use of the symmetry axiom (of the interface \code{Set}). \

@@

The operations, constants and properties of each algebraic structure will be described in a dedicated interface. The first structure, completely trivial, is the magma. A magma is a structure built on top of \code{Set}, that just adds a \code{Plus} operation, and no specific properties about it :

@@

This code means that a type $c$ (for $carrier$) is a \code{Magma} if it is already a \code{Set} (ie, it is equipped with the equivalence relation $\simeq$ and the equivalence test \code{set\_eq}), and if it has a $+$ operation.
In fact, there is an additional requirement that will apply for all operators (in this case, the $+$ operation), which is that they need to be "compatible" with the equivalence relation, which is expressed by the following axiom for $+$ : \\ $\code{Plus\_preserves\_equiv} : \{c1:c\} \rightarrow \{c2:c\} \rightarrow \{c1':c\} \rightarrow \{c2':c\} \rightarrow (c1 \simeq c1') \rightarrow (c2 \simeq c2') \rightarrow ((c1+c2) \simeq (c1'+c2'))$ \\ 
This requirement comes from the fact that we're dealing with any equivalence relation. The user is free to use his own equivalence relation, but it should be compatible with the operations that he is using.
As it was the case for \code{Set}, there is no tactic to write for \code{Magma}, because there is no property at all : all equivalences are again syntactic equivalences, and they can all be proven by \code{refl}.

A semi-group is a magma (ie, it still has a \code{Plus} operation), but moreover it has the property of associativity for this operation.

@@

As example of magma, we can cite \code{Nat} equipped with its operation of addition and \code{List} and the operation of concatenation.

The Monoid structure is a Semi-Group with the property of neutral element for a distinguished element called \code{Zero}.

@@

The hierarchy of interfaces continues with Commutative-Monoid, Group, Commutative Group, Semi-Ring and Ring, but we only show the \code{Group} interface for reasons of space :

@@

The notion of group uses two new operations (\code{Neg} and \code{Minus}), but \code{Minus} is not primitive and can be simplified with $+$ and \code{Neg}. The important property of a group is that every element \code{c1} must admit \code{Neg\ c1} as inverse element for the $+$ operation.

@@

When trying to prove an equivalence $x \simeq y$, the universally-quantified variables have been abstracted, and they are now part of the context. Our tatics will work by normalising both sides of the "potential equivalence" $x \simeq y$, and afterwards comparing them with a syntactic equality test. The problem is that such a normalisation function will need to do different treatments for a a variable, for a constant and for an operator. For this reason, we will need to work by reflection. There will be one type of reflected terms for each algebraic structure. The novelty is not the use of reflected terms (this has already been done, as discussed in section~\ref{sect:relatedWork}), but is the use of a type-safe reflection mechanism where we index the reflected terms by the concrete value that they represent. Each of these datatype is parametrised over a type \code{c}, which is the real type on which we want to prove equalities (the \emph{carrier} type). It is also indexed over an implementation of the corresponding interface for \code{c} (we usually call it \code{p}, because it behaves as a \emph{proof} telling that the structure \code{c} has the desired properties), indexed over a context of abstracted variables (a vector $\Gamma$ of \code{n} elements of type \code{c}), and most important it is also indexed over a value of type \code{c}, which is precisely the concrete value being encoded.

A magma is only equipped with the operation of addition. Thus, we only have three concepts to express in order to reflect terms in a Magma : constants, variables, and additions.

@@

For an expression $e_x\ :\ \code{ExprMa}\ \Gamma\ x$, we will say that $e_x$ denotes (or encodes) the number $x$ in the context $\Gamma$.
When an expression is a variable \code{(VarMa\ \_\ $\Gamma$\ i)}, the denoted number is simply the corresponding variable in the context, ie, \code{(index\ i\ $\Gamma$)}.
The expression \code{(ConstMa\ \_\ $\Gamma$\ k)} denotes the constant $k$ of type $c$. Finally, if $e_x$ is an expression encoding the concrete value $x$, and $e_y$ is an expression encoding the concrete value $y$, then the expression \code{PlusMa\ $e_x$\ $e_y$} denotes the concrete value $(x + y)$. Because the reflected terms embed their corresponding inputs, they are guarantee to be faithful representations of them.

There is no additional operations in SemiGroup and Monoid, so the datatypes that reflect terms in these two structures will have exactly the same shape as the one for Magma that we have given above.
However, the \code{Group} interface will have to introduce two new constructors for the \code{Neg} and \code{Minus} operations :

@@

The index of type $c$ (the real value encoded by an expression) is always expressed by using the lookup function \code{index} and the available operations in the implementation $p$, which for a group are $+$, \code{Minus} and \code{Neg}.

@@

We want to write the normalisation function on a \emph{correct by construction} way, which means that no additional proof should be required after the definition of the function. Thus, \code{norm} will produce the proof that the new expression freshly produced has the same interpretation as the original expression, and this will be made easier by the fact that all the datatypes for reflecting terms (\code{ExprMa}, \code{ExprG}, \code{ExprR}, etc) are indexed over the real --concrete-- value : a term of type $\code{Expr}\ \Gamma\ x$ is the encoding of the concrete value $x$ in the context $\Gamma$.
Thus, for each algebraic structure, the type of the function \code{norm} will look like this : \\
$\code{norm} :\ \code{Expr}\ \Gamma\ x\ \rightarrow\ (x'\ \code{**}\ (\code{Expr}\ \Gamma\ x',\ x\ \simeq\ x'))$ \\
Every function of normalisation \code{norm} will produce a dependent pair : the new concrete value $x'$, and a pair made of an $\code{Expr}\ \Gamma\ x'$ which is the new encoded term indexed over the new concrete number we have just produced, and a proof that old and new concrete values $x$ and $x'$ are equivalent.

This function doesn't simply produce an $\code{Expr}\ \Gamma\ x$ (with an internal conversion of type) because the proof of $x \simeq x'$ is an essential component for building the desired proof of $x \simeq y$, which is the main problem that we are trying to solve.

We will explain how the tactic works by commenting the construction of the tactic for Group, but the other algebraic structures will be equipped similarly. The equivalence we are trying to prove is $x \simeq y$, where $x$ and $y$ are elements of the type $c$, which  simulates a group. The reflected term for $x$ will be denoted $e_x$, and this term will have the type $\code{ExprG}\ p\ \Gamma\ x$, which means that $e_x$ is guaranteed to be the encoding of $x$ because $e_x$ is precisely indexed over the real value $x$.
We've got similar things for $y$, which is encoded by $e_y$, and its type is indexed over the real value $y$.
Running the normalisation procedure on $e_x$ will produce the normal form $e_{x'}$ of type $\code{ExprG}\ p\ \Gamma\ x'$ and a proof $p_x$ of $x \simeq x'$ while running the normalisation procedure on $e_y$ will produce the normal form $e_{y'}$ of type $\code{ExprG}\ p\ \Gamma\ y'$ and a proof $p_y$ of $y \simeq y'$.
It is now enough to simply compare $e_{x'}$ and $e_{y'}$ using a standard syntactic equality test \code{exprG\_eq} because these two expressions are now supposed to be in normal form.

### 2017-03-30 paper-CICM-17/paper.tex
paper CICM-17 : fixed subtitle
some fixes

@@

\mainmatter              % start of the contributions
%
\title{Automatically Proving Equivalence by Type-Safe Reflection}
\subtitle{Or how to use dependent types to solve some of the problems they have caused}
%
\titlerunning{xxx}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Franck Slama \and Edwin Brady}
\institute{University of St Andrews, Scotland, UK,\\
\email{fs39@st-andrews.ac.uk, ecb10@st-andrews.ac.uk}}

>>>>

\def\code#1{{\mbox{\texttt{#1}}}} % For short element of code in the test that should not be hyphenated
\def\codeL#1{{\texttt{#1}}} % For Long element of code in the text that might need to be hyphenated

@@

\mainmatter              % start of the contributions
%
\title{Automatically Proving Equivalence by Type-Safe Reflection}
\subtitle{Or how to use dependent types to address some of the problems they have caused}
%
\titlerunning{xxx}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Franck Slama \and Edwin Brady}
\institute{University of St Andrews, Scotland, UK,\\
\email{fs39@st-andrews.ac.uk, ecb10@st-andrews.ac.uk}}
